---
title: "Group analysis"
output: html_notebook
---
```{r}
library(readxl)
library(tidyverse)
library(vegan)
library(ggplot2)
library(multcompView) #tukey?
library(RVAideMemoire) #pairwise manova
library(ape)
library(ggtree)
library(mvMORPH)
library(colorspace)
```

```{r}
me_matrix <- read_excel(path = "Methyl Ether Matrix.xlsx", sheet = "Methyl Ether Matrix (FINAL)")
me_matrix <- as.data.frame(me_matrix)
me_names <- me_matrix[1,] #get colnames
me_matrix <-  me_matrix %>%
              filter(!is.na(me_matrix[,1])) #remove unapproved files

me_matrix <- me_matrix[-1,] #remove extra row from the top
row.names(me_matrix) <- me_matrix$...2 #assign row names to file names

me_matrix <- select(me_matrix, !which(is.na(me_names))) #keep only named columns
colnames(me_matrix) <- me_names[!is.na(me_names)] #assign column names to ME names
me_matrix <- select(me_matrix, !"Unidentified Methyl Ethers?") #remove unidentified column

me_matrix <- apply(me_matrix, c(1,2), as.numeric) #convert from character vectors to numeric vectors
me_matrix <- as.data.frame(me_matrix) #convert back to data frame

me_matrix <- me_matrix %>%
              #filter(grepl("_F_", rownames(me_matrix))) %>% #select only M or F
              apply(2, replace_na, replace=0) %>% #replace nas
              as.data.frame() %>% #back to df
              select(-1) %>% #remove C12
              filter_all(any_vars(.> 0)) #remove rows without any MEs (i.e., keep rows where abundance of any ME >0)

#Normalize the data (divide each cell by sum of the whole row)
me_matrix <- me_matrix/rowSums(me_matrix)

#get location and other info from Mass Spectral Identification Progress
#manual processing: replace the misidentified kama and rest
#fix some of the dates

progress_sheet <- read_excel(path = "Mass Spectral Identification Progress.xlsx", sheet = "Dragline DCM (FINAL)")
progress_sheet <- as.data.frame(progress_sheet)
cols <- progress_sheet[1,]
rows <- progress_sheet$`...1`
progress_sheet <- progress_sheet[-1,] %>% #remove top row
                  select(-1) #remove first column
rownames(progress_sheet) <- rows[2:length(rows)]
colnames(progress_sheet) <- cols[2:length(cols)]

me_matrix_info <- merge(progress_sheet, me_matrix, by = 0) #merging by file name
me_matrix_info$Species <- as.factor(me_matrix_info$Species) #converting to factors in case we want to group
me_matrix_info$Sex <- as.factor(me_matrix_info$Sex)
me_matrix_info$`Collection Location` <- as.factor(me_matrix_info$`Collection Location`)
me_matrix_info$`Spider Type` <- as.factor(me_matrix_info$`Spider Type`)

rownames(me_matrix_info) <- me_matrix_info$Row.names
me_matrix_info <- me_matrix_info %>%
  select(-1)
nrow(me_matrix_info) #should be 246
```
Here, subset the full data frame. Select the only files which have data for the grouping variable you are considering, select only females/males, etc.
```{r}
selected_matrix <- me_matrix_info %>%
  #filter(`Species` == "Waika") %>%
  #filter(`Collection Location` != "?") %>% 
  #filter(`Collection Location` == "Lower Waikamoi") %>% 
  #filter(Species != "Kama") %>%
  #filter(Species != "Waika") %>%
  #filter(Species != "Quasi") %>%
  #filter(Species != "Acuta") %>%
  #filter(Species != "Tri") %>%
  #filter(Species != "Eury") %>%
  filter(!is.na(`Species`)) #replace "Collection Location" with whichever division you're considering (species, type, etc.)

selected_me_matrix <- me_matrix %>%
  filter(row.names(me_matrix) %in% row.names(selected_matrix)) 
```

Select the grouping variable you're considering.
```{r}
grouping_var <- selected_matrix$`Species`
```

PERMANOVA, PERMDISP, Pairwise PERMANOVA, Tukey test using Bray-Curtis distance
```{r}
print("ADONIS results:")
adonis(selected_matrix %>% select(starts_with("ME ")) ~ grouping_var, method="bray", perm=999) 
```

```{r}
#betadisper
print("betadisper results:")
betadisp <- betadisper(vegdist(selected_matrix %>% select(starts_with("ME ")), method = "bray"), grouping_var)
betadisp
```

```{r}
#permutest
print("permutest results:")
permutest(betadisp, pairwise = FALSE,
          permutations = 999)
```

```{r}
pair_manova <- pairwise.perm.manova(vegdist(selected_me_matrix, method = "bray"), grouping_var,nperm=999)
pair_manova
```

```{r}
#tukey test, where the one variable being considered is mean distance to centroid
TUKEY <- TukeyHSD(betadisp, conf.level=0.95)
par(mar=c(5,6,4,1)+.1)
plot(TUKEY , las=1 , col="brown")
```
```{r}
TUKEY
```
Calculating and visualizing bouquet complexity (defined as number of compounds present)

```{r}
#measuring bouquet complexity by number of compounds present
#add a column which counts the number of non-zero areas for methyl ethers in each file
selected_me_matrix_info <- selected_matrix
selected_me_matrix_info <- cbind(selected_me_matrix_info, "Num_MEs" = apply(selected_me_matrix, 1, function (x) sum(ifelse(x!=0, 1, 0))))

#boxplot grouped by Spider Type (could be Species, or Location -- any grouping variable)
ggplot(data = selected_me_matrix_info, aes(x= `Spider Type`, y = Num_MEs, fill = `Spider Type`)) + 
  geom_boxplot() + 
  theme_minimal() +
  ggtitle("All")
```

Calculating average pairwise dissimilarity
```{r}
#all files, conspecific and interspecific distances
d <- vegdist(selected_matrix %>% select(starts_with("ME ")), method = "bray") #create a distance matrix based on the methyl ether matrix

avg_dist_all <- mean(as.vector(t(d)))
sd_dist_all <- sd(as.vector(t(d)))

#species wise pairwise distances
d_df <- as.data.frame(as.matrix(d))
avg_dist_species <- c()
sd_dist_species <- c()
cons_dist <- c() #conspecific distances
inter_dist <- c() #interspecific distances

for (species in levels(me_matrix_info$Species)){
  species_d <- d_df %>% #conspecific data
    filter(grepl(species, rownames(d_df))) %>% #filter only rows of the species
    select(starts_with(species)) %>% #select only columns of the species
    as.dist() #convert back to distance matrix
  species_dist_vec <- as.vector(t(species_d))
  names(species_dist_vec) <- rep(species, times = length(species_dist_vec))
  avg_dist_species <- c(avg_dist_species, mean(species_dist_vec))
  sd_dist_species <- c(sd_dist_species, sd(species_dist_vec))
  cons_dist <- c(cons_dist, species_dist_vec)
  
  #interspecific data
  inter_d <- d_df %>% 
    filter(grepl(species, rownames(d_df))) %>%  #filter rows of the species
    select(!starts_with(species)) #select columns that are not of the species
  inter_dist_vec <- as.vector(t(inter_d)) 
  names(inter_dist_vec) <- rep("Interspecies", times = length(inter_dist_vec))
  inter_dist <- c(inter_dist, inter_dist_vec) 
  
  #subsetting distance matrix to avoid double counting interspecific
  d_df <- d_df %>%
    filter(!grepl(species, rownames(d_df))) %>% #remove rows of the species
    select(!starts_with(species)) #remove columns of the species
}

names(avg_dist_species) <- levels(me_matrix_info$Species)
names(sd_dist_species) <- levels(me_matrix_info$Species)

avg_dist_cons <- mean(cons_dist, na.rm = TRUE)
sd_dist_cons <- sd(cons_dist, na.rm = TRUE)

avg_dist_inter <- mean(inter_dist, na.rm = TRUE)
sd_dist_inter <- sd(inter_dist, na.rm = TRUE)

dist_data <- data.frame("Distance" = c(cons_dist, inter_dist), "Species" = c(names(cons_dist), names(inter_dist)))
dist_data$inter <- ifelse(dist_data$Species == "Interspecies", "Inter", "Cons")
```

```{r}
print("Average pairwise Bray-Curtis distances")
avg_dist_all
print("Std. Dev of pairwise Bray-Curtis distances")
sd_dist_all
```

```{r}
print("Average pairwise Bray-Curtis distances between all conspecific pairs")
avg_dist_cons
print("Std. Dev of pairwise Bray-Curtis distances between all conspecific pairs")
sd_dist_cons
```
```{r}
print("Average pairwise Bray-Curtis distances between conspecific pairs, for each species")
avg_dist_species
print("Std. Dev of pairwise Bray-Curtis distances between conspecific pairs, for each species")
sd_dist_species
```
```{r}
print("Average pairwise Bray-Curtis distance between all interspecific pairs")
avg_dist_inter
print("Std. Dev of pairwise Bray-Curtis distance between all interspecific pairs")
sd_dist_inter
```
T-test comparing interspecific and conspecific pairwise Bray-Curtis distances
```{r}
dist_data <- data.frame("Distance" = c(cons_dist, inter_dist), "Species" = c(names(cons_dist), names(inter_dist)))
dist_data$inter <- ifelse(dist_data$Species == "Interspecies", "Inter", "Cons")
t.test(dist_data$Distance ~ dist_data$inter)
wilcox.test(dist_data$Distance ~ dist_data$inter)
```
```{r}
#box and whisker plots of dissimilarity
ggplot(data = dist_data, aes(x=Species, y = Distance, fill = Species)) + 
  geom_boxplot() + 
  theme_minimal() +
  ggtitle("All")
```

Plotting avg distance vs no. of samples (check if distance is a function of the no. of samples)
```{r}
counts <- me_matrix_info %>% count(Species)
counts <- counts$n
count_data <- data.frame(counts = counts, avg_dist = avg_dist_species)
plot(x = count_data$counts, y = count_data$avg_dist, type = "p")
linear_mod <- lm(avg_dist ~ counts, count_data)
abline(lm(avg_dist ~ counts, count_data))
```
```{r}
summary(linear_mod)
```

Heatmap (do we need this??)
```{r}
species_me_matrix <- aggregate(me_matrix_info %>% select(starts_with("ME ")), 
          by = list(Species = me_matrix_info$Species), 
          FUN = mean)
species_me_matrix <- select(species_me_matrix, -1)
collapsed_tree_without_palu <- read.tree(file = "phylogeny_pruned_without_palu")
rownames(species_me_matrix) <- sort(collapsed_tree_without_palu$tip.label)
colnames(species_me_matrix) <- substr(colnames(species_me_matrix), 4, 7)
p <- ggtree(collapsed_tree_without_palu) + geom_tiplab()
gheatmap(p, species_me_matrix, 
         low = "lightcyan", high = "navy", color = "lightcyan", 
         colnames_angle = 90, font.size = 3,
         width = 4, offset = 0.04)
```

